A conditional permutation-based approach to test confounder effect and center-bias in machine learning models

### 2021-08-16 manuscript.tex
small changes

The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via "restricted permutations" that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature) However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances. Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control with biased classification problems.
The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, by design, does not provide any quantification of statistical significance.
Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the the predictors are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both. It's computation assumes 'faithfulness' (if two variables are independent, there is no direct influence between the two in the underlying graph) a strong condition that might not be fulfilled in many real applications (e.g. functional MRI is an \emph{indirect} measure of neural activation, rendering the alternative hypothesis of this test true in absence of any real confounder).

@@

Analysis code is available as jupyter notebook\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb} in the github repository of the 'mlconfound' package.

@@

Tangent correlation across the timeseries of the n=122 regions of the BASC \citep{bellec2010multi} atlas) was computed with nilearn\footnote{http://nilearn.github.io/} \citep{huntenburg2017loading, esteve2015big}.

@@

Analysis code is available as jupyter notebook\footnote{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb} in the github repository of the 'mlconfound' package.

>>>>

The permutation-based approach of \cite{chaibub2019permutation} evaluates the null-hypothesis of no confounder bias via "restricted permutations" that retain the observed ratios of class labels within batches (often referred to as balanced permutations in the literature) However, as shown by \cite{southworth2009properties} and further discussed by \cite{hemerik2018exact}, balanced permutations do not provide a correct null behavior and can be invalid in certain circumstances. Indeed, \cite{ferrari2020measuring} demonstrated that this method does not guarantee a proper type I error control with biased classification problems.
The 'confounder index' (CI) proposed by \cite{ferrari2020measuring}, by design, does not provide any quantification of statistical significance.
Moreover, both approaches involve repeated re-fitting of the machine learning model, which might not be feasible for models with high computational complexity (especially when hyperparameters are optimized in a nested cross-validation).
The Kolmogorov Complexity-based causal inference test, as proposed by \cite{wachinger2021detect}, aims at determining whether it is more likely that the predictors are a direct cause of the target variable or, alternatively, there exists an unobserved random variable that is the cause of both. It's computation assumes 'faithfulness' (if two variables are independent, there is no direct influence between the two in the underlying graph) a strong condition that might not be fulfilled in many real applications (e.g. functional MRI is an \emph{indirect} measure of neural activation, rendering the alternative hypothesis of this test true in absence of any real confounder).

@@

Analysis code is available as jupyter notebook\footnote{\href{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb}{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_hcp.ipynb}} in the github repository of the 'mlconfound' package.

@@

Tangent correlation across the timeseries of the n=122 regions of the BASC \citep{bellec2010multi} atlas) was computed with nilearn\footnote{\href{http://nilearn.github.io/}{http://nilearn.github.io/}} \citep{huntenburg2017loading, esteve2015big}.

@@

Analysis code is available as jupyter notebook\footnote{\href{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb}{https://github.com/pni-lab/mlconfound/tree/master/notebooks/analysis\_abide.ipynb}} in the github repository of the 'mlconfound' package.
